---
title: "Time series of stochastic process variables"
subtitle: "Multi-agent system for smart machining, part II"
author: "Paolo Bosetti"
format: revealjs
---


```{r setup}
options(width = 60)
set.seed(0)
library(tidyverse)
library(xts)
library(lubridate)
library(tsbox)
library(ggfortify)
library(ggpubr)
theme_set(theme_gray()+theme(legend.position = "bottom"))

sample_time <- ymd_hms("2026-03-09 08:00:00", tz = "UTC") +
  c(0, 5, 8, 15, 23, 31) * 60
sample_value <- c(20.0, 20.4, 20.1, 20.8, 21.2, 21.0)
sensor_xts <- xts(sample_value, order.by = sample_time)
sensor_df <- tibble(timestamp = sample_time, value = sample_value)
```

# Time and timestamps

In a time series, each observation is paired with a **time coordinate** that gives order, spacing, and physical meaning to the data

The same physical instant can be stored in different ways: readable text, numeric offset from an epoch, or a native date-time object provided by the language or the operating system

Choosing the wrong representation can create ambiguities in time zone, resolution, clock drift, or query semantics

## Storing time

Two common representations are used in practice:

* **ISO8601 date-time string**, for example `2026-03-09T08:15:30.125Z`
* **Numeric offset from an epoch**, for example seconds or milliseconds from a reference instant

ISO strings are readable, self-describing, and easy to exchange between systems, especially when they include time zone information

Epoch-based values are compact and fast for storage, sorting, and arithmetic, but they require agreement on the epoch, units, and time zone convention

In industrial applications, timestamps are often stored with **millisecond resolution**, although sensor acquisition may internally use finer clocks

## Epoch conventions and clocks

:::columns
:::column
An **epoch** is the reference instant from which a numeric timestamp is measured

Common conventions include:

* Unix epoch: `1970-01-01 00:00:00 UTC`
* GPS epoch: `1980-01-06 00:00:00 UTC`
* Other software-specific origins, used by spreadsheets, PLCs, or proprietary databases

If the epoch changes, the same numeric value refers to a different physical instant
:::

:::column
Operating systems expose different clocks:

* **System clock**: returns civil time and can jump because of NTP, manual updates, or daylight-saving adjustments
* **Monotonic clock**: only increases and is used to measure elapsed time reliably

Therefore:

* use the **system clock** for timestamps attached to data
* use a **monotonic clock** for latency, timeout, and sampling-period measurements
:::
:::

:::aside
Never compute elapsed time by subtracting two readings of the wall-clock if the clock can be corrected while the program is running.
:::

## Monotonic clock limitations

The **monotonic clock** is usually referenced to an internal origin that depends on the current operating-system session, typically the **last boot time**

Therefore, its zero is not a civil date and is generally different on each machine, even if the devices are nominally synchronized on the network

As a consequence, monotonic time is excellent for measuring:

* elapsed time
* latency
* timeout intervals
* local acquisition jitter

However, it **cannot be used to synchronize measurements collected on different devices**, because the timestamps do not share a common physical epoch

To align data acquired by multiple systems, one needs timestamps derived from a shared wall-clock reference, typically UTC, distributed for example by NTP, PTP, or GNSS

## Time in R

R provides several complementary tools for time handling:

* `Date`: stores calendar dates without time-of-day
* `POSIXct`: stores date-time as seconds from the Unix epoch
* `ts`: for **regularly sampled** series with known frequency
* `xts`: for indexed time series, especially convenient for date-time filtering and irregular sampling
* `lubridate`: simplifies parsing, rounding, shifting, and extracting date-time components

The choice depends on whether time is regular, irregular, or mainly used for querying

## Regular series with `ts`

:::columns
:::column
The `ts` class is designed for **regular sampling**, such as monthly production, daily energy use, or hourly average demand

It does not store a full timestamp per sample. Instead, it stores:

* a start instant
* a frequency
* the ordered values

This is efficient and works very well with classical time-series models
:::

:::column
```{r}
ap <- AirPassengers
autoplot(ap) +
  labs(x = "Year", y = "Passengers",
       title = "Monthly series stored with ts")
```
:::
:::

## Indexed series with `xts`

:::columns
:::column
The `xts` class is more convenient when timestamps are explicit and may be irregular

Each observation is paired with a time index, typically `Date` or `POSIXct`

This is useful for:

* sensor logs
* event streams
* asynchronous acquisition
* filtering by calendar date or time window
:::

:::column

:::panel-tabset
## Plot
```{r}
autoplot(sensor_xts) +
  labs(x = "Timestamp", y = "Value",
       title = "An xts object with explicit timestamps")
```

## Data
```{r}
sensor_xts %>% 
  fortify() %>%
  knitr::kable(row.names = TRUE)
```
:::
:::
:::

## Querying `xts` with date ranges

:::columns
:::{.column width="45%"}
One of the strengths of `xts` is its compact indexing syntax

Examples:

* one day: `x["2026-03-09"]`
* one range: `x["2026-03-09/2026-03-10"]`
* from an instant onward: `x["2026-03-09 08:10:00/"]`
* up to an instant: `x["/2026-03-09 08:15:00"]`
:::

:::{.column width="55%"}
```{r echo=TRUE}
sensor_xts["2026-03-09 08:05:00/2026-03-09 08:20:00"]
```

```{r echo=TRUE}
sensor_xts["/2026-03-09 08:08:00"]
```
:::
:::

This is often faster to read and write than manual logical filtering

## `lubridate` for parsing and filtering

:::columns
:::column
`lubridate` helps create robust pipelines for date-time handling

Typical tasks include:

* parsing strings with `ymd_hms()`
* changing time zone with `with_tz()` or `force_tz()`
* rounding with `floor_date()` and `ceiling_date()`
* extracting fields such as `year()`, `month()`, `wday()`, `hour()`
:::

:::column
```{r echo=TRUE}
sensor_df %>%
  mutate(
    minute_bin = floor_date(timestamp, "10 min"),
    hour = hour(timestamp))
```

```{r echo=TRUE}
sensor_df %>% filter(
  timestamp >= ymd_hms("2026-03-09 08:05:00"),
  timestamp <  ymd_hms("2026-03-09 08:25:00"))
```
:::
:::

:::{.callout-note}
This is especially useful before converting tabular data to `xts`
:::

## From data frame to `xts`

To work efficiently with time series in R, a common workflow is:

::: {.columns}
::: {.column}
1. import tabular data
2. parse the `timestamp` column as `POSIXct`
3. sort by time
4. convert to `xts`
5. query and aggregate
:::
::: {.column}
```{r echo=TRUE}
sensor_tbl <- tibble(
  timestamp = c("2026-03-09 08:00:00",
                "2026-03-09 08:05:00",
                "2026-03-09 08:08:00"),
  temperature = c(20.0, 20.4, 20.1)
)

sensor_tbl2 <- sensor_tbl %>%
  mutate(timestamp = ymd_hms(timestamp)) %>%
  arrange(timestamp)

sensor_xts2 <- xts::xts(
  x = sensor_tbl2$temperature,
  order.by = sensor_tbl2$timestamp
)

sensor_xts2
```
:::
:::
<!-- end columns -->

## Time series in Python

Python offers libraries with roles similar to those seen in R:

* `datetime`: basic date-time objects, time zones, timedeltas
* `pandas`: indexed time series, resampling, rolling windows, filtering by date range
* `numpy`: efficient numerical arrays and vectorized operations
* `statsmodels`: AR, MA, ARIMA, seasonal models, ACF and PACF
* `polars`: fast tabular processing, useful before conversion to time-indexed data
* `xarray`: labeled multi-dimensional arrays, useful for sensor grids and scientific data

In practice, `pandas` plays a role similar to a combination of `data.frame`, `xts`, and part of `lubridate`

## Time series in Python

:::columns
:::column
Typical workflow in Python:

* parse timestamps with `pd.to_datetime()`
* set the timestamp as index
* sort the index
* slice by date or time range
* resample or aggregate if needed

This is very close to the `xts` workflow in R
:::

:::column
```python
import pandas as pd

df = pd.DataFrame({
    "timestamp": [
        "2026-03-09T08:00:00Z",
        "2026-03-09T08:05:00Z",
        "2026-03-09T08:08:00Z"
    ],
    "value": [20.0, 20.4, 20.1]
})

df["timestamp"] = pd.to_datetime(df["timestamp"], utc=True)
df = df.sort_values("timestamp")
df = df.set_index("timestamp")

window = df.loc["2026-03-09 08:00:00":"2026-03-09 08:06:00"]
```
:::
:::

## R and Python correspondence

:::columns
:::column
Useful parallels are:

* `ts` in R and `statsmodels` regular-series tools in Python
* `xts` in R and `pandas` with `DatetimeIndex` in Python
* `lubridate` in R and `pandas` date-time accessors, plus `datetime`, in Python
:::

:::column
```python
# resample every 10 minutes
agg = df.resample("10min").mean()

# rolling mean
smooth = df["value"].rolling(3).mean()

# month and hour extraction
df["hour"] = df.index.hour
df["day"] = df.index.day
```
:::
:::

::: {.callout-note}
The conceptual workflow is almost the same:

parse {{< bi caret-right-fill >}} index {{< bi caret-right-fill >}} filter {{< bi caret-right-fill >}} resample {{< bi caret-right-fill >}} model
:::

# Time-series queries in MongoDB

A common document structure for process data in MongoDB is:

```javascript
{
  machine_id: "M01",
  timestamp: ISODate("2026-03-09T08:15:30.125Z"),
  temperature: 20.8,
  vibration: 0.031
}
```

The key point is that all documents share a `timestamp` field stored as an **ISO date**, typically with millisecond resolution

## Querying a time range in MongoDB

::: {.columns}
::: {.column}
```json
{
  "_id": {
    "$oid": "698afb51c3b4084d840b9c85"
  },
  "timestamp": {
    "$date": "2026-02-10T09:33:05.344Z"
  },
  "measure": {
    "back_bend": -0.5168184041976929,
    "back_flex": -1.20260488986969,
    "back_rot": 1.749389410018921,
    "cervical_flex": 17.3467960357666,
    "horiz_reach_left": 0.20939064025878906,
    "horiz_reach_right": 0.20467370748519897,
    "hostname": "nuc3",
    "stability_margin": 0.11027171462774277,
    "timecode": 37985.2,
    "vert_reach_left": 14.016213417053223,
    "vert_reach_right": 10.769205093383789
  }
}
```
:::
::: {.column}
The standard pattern is:

* inclusive lower bound: `$gte`
* exclusive upper bound: `$lt`

This avoids overlap when adjacent windows are queried one after another

```javascript
db.measurements.find({
  timestamp: {
    $gte: ISODate("2026-02-10T08:00:00.000Z"),
    $lt:  ISODate("2026-02-10T10:00:00.000Z")
  }
}).sort({ timestamp: 1 })
```
:::
:::
<!-- end columns -->


:::aside
For performance, a range query on `timestamp` should be supported by an index on the same field, or by a compound index such as `{ machine_id: 1, timestamp: 1 }`.
:::

## MongoDB query in JavaScript

```javascript
const t0 = ISODate("2026-03-09T08:00:00.000Z");
const t1 = ISODate("2026-03-09T09:00:00.000Z");

db.measurements.find(
  {
    machine_id: "M01",
    timestamp: { $gte: t0, $lt: t1 }
  },
  {
    _id: 0,
    machine_id: 1,
    timestamp: 1,
    temperature: 1,
    vibration: 1
  }
).sort({ timestamp: 1 });
```

## MongoDB query in Python

```python
from datetime import datetime, timezone
from pymongo import MongoClient
client = MongoClient("mongodb://localhost:27017")
db = client["plant"]
col = db["measurements"]
t0 = datetime(2026, 3, 9, 8, 0, 0, tzinfo=timezone.utc)
t1 = datetime(2026, 3, 9, 9, 0, 0, tzinfo=timezone.utc)
cursor = col.find(
    {
        "machine_id": "M01",
        "timestamp": {"$gte": t0, "$lt": t1}
    },
    {
        "_id": 0,
        "machine_id": 1,
        "timestamp": 1,
        "temperature": 1,
        "vibration": 1
    }
).sort("timestamp", 1)
rows = list(cursor)
```

## MongoDB query in R

```r
library(mongolite)
library(glue)
m <- mongo(collection = "measurements", db = "plant", url = "mongodb://localhost:27017")
t0 <- "2026-03-09T08:00:00.000Z"
t1 <- "2026-03-09T09:00:00.000Z"
query <- glue(
'{{
  "machine_id": "M01",
  "timestamp": {{
    "$gte": {{"$date": "{t0}"}},
    "$lt":  {{"$date": "{t1}"}}
  }}
}}'
)
fields <- '{"_id": 0, "machine_id": 1,
            "timestamp": 1, "temperature": 1,
            "vibration": 1}'
df <- m$find(query = query, fields = fields, sort = '{"timestamp": 1}')
```

## From MongoDB query to time series

Once the query result is available, the workflow is again the same:

* parse the returned `timestamp`
* sort rows by time
* convert to a time-indexed object
* analyze or visualize the selected window

```{r}
#| echo: TRUE
#| output-location: column
#| fig-height: 2
sensor_df %>%
  arrange(timestamp) %>%
  ggplot(aes(x = timestamp, y = value)) +
  geom_line() +
  geom_point() +
  labs(x = "Timestamp", y = "Value",
       title = "Selected time window")
```

## Summary

* Time can be stored as readable ISO text or as numeric offset from an epoch
* Epoch convention, time zone, and clock type must be known to avoid semantic errors
* In R, `ts` is convenient for regular series, while `xts` is preferable for explicit timestamp indexing
* In Python, `pandas` provides the closest workflow to `xts` plus data-frame filtering
* MongoDB range queries on ISO dates are naturally expressed with `$gte` and `$lt`
* Once data are extracted, the main tasks are always parse, index, filter, aggregate, and model
