---
title: "Time-domain analysis of time series"
subtitle: "Multi-agent system for smart machining, part II"
author: "Paolo Bosetti"
format: revealjs
---


```{r setup}
options(width = 60)
set.seed(0)
library(latex2exp)
library(ggpubr)
library(glue)
library(tidyverse)
library(ggfortify)
library(modelr)
library(tsbox)
library(xts)
library(astsa)
library(patchwork)
theme_set(theme_gray()+theme(legend.position = "bottom"))
```

# Time Series

A **time series** consists of a sequence of observations of a random variable such that the influence of one observation on the following ones cannot be neglected and, therefore, dependence on time becomes essential

## Time-series statistics

All regression methods seen so far are based on the assumption that the random variable is $x\overset{IID}\sim \mathcal{N}(\mu, \sigma^2)$. That is, all observations **must be uncorrelated**

This assumption is, among other things, the basis for the recommendation to randomize the operating sequence

Suppose we can treat a measurement as a time-dependent signal. It is clear that by reducing the signal sampling interval, sooner or later each sample will be correlated with the previous one

There is therefore a **maximum sampling frequency** above which every measurement turns out to be autocorrelated, that is, **the observations of $x$ are no longer IID**

This situation occurs when the **dynamics** of the measuring instrument or of the measurand itself, which are **always finite**, are slower than the time interval at which measurements are taken

## Time-series statistics

:::columns
:::column
Consider, for example, the time series in the figure, which reports the difference between the mean temperature of land areas and the corresponding mean value over the period 1951--1980

It is clear that nearby observations are more correlated than distant ones

Moreover, it is clear, and relevant, to evaluate the dependence of the random variable on time in order to **make future predictions**

:::
:::column
```{r}
ts <- ts_xts(gtemp_land)
ref <- ts["1951/1980"]
ggplot(ts, aes(x=Index, y=value)) + 
  geom_line(color=gray(0.5)) + 
  geom_line(data=ref, aes(y=value, color="Reference Period"), linewidth=1) +
  # geom_point(shape=4) +
  geom_hline(yintercept=0, linetype=2) +
  scale_x_date(breaks="20 years", date_labels = "%Y") +
  labs(x="Year", y="Land Temperature Anomaly (Â°C)", color="") +
  theme(legend.position = "bottom")
```

:::
:::


## Autocovariance and autocorrelation

We have seen how the **covariance** and **correlation** operators are used to estimate the independence of two samples

Considering a time-dependent signal $x=x(t)$, it is interesting to consider the covariance of the signal with itself, shifted in time

We define the **autocovariance function** $\gamma(s,t)$ as the function that evaluates the covariance of a time signal with itself evaluated **starting** at times $s$ and $t$:
$$
\gamma_x(s, t) = \sigma_{x_s, x_t} = E[(x_s-\mu_s)(x_t-\mu_t)]
$$

It is clear that $\gamma_x(s,s)=\sigma^2(x_s)$

The **autocorrelation function** (ACF), consequently, is defined as:
$$
\rho_x(s, t) = \frac{\gamma_x(s,t)}{\sqrt{\gamma_x(s,s)\gamma_x(t,t)}}
$$
and has the advantage of always lying in $[-1,1]$. It is also clear that $\rho_x(s,s)=1$




## Autocorrelation
If we sample a continuous signal at fixed intervals $\Delta t$ over a total duration $T$, we obtain a **finite time series** of $N=T/\Delta t$ observations: $x_{0}=\left<x_1, x_2, \dots, x_N\right>$

We can extend the definition by setting $s=t_0 = 0$ as the initial instant of the series and $t=s+\tau$ as a generic later instant such that
$$
\tau = \ell \Delta t
$$
where $\ell$ is the delay or *lag*, and then the **autocovariance** and **autocorrelation for a finite time series** become:
$$
\gamma_x(\ell) = \frac{\sum_{i=1}^{N-\ell} (x_i - \bar x_0) (x_{i+\ell}-\bar x_\ell) }{N-\ell-2},\quad \rho_x(\ell) = \frac{\gamma_x(\ell)}{\sqrt{\sigma_x(0)\sigma_x(\ell)}}
$$
where
$$
\bar x_0=\frac{1}{N-\ell}\sum\nolimits_{i=1}^{N-\ell} x_i,\quad \bar x_\ell=\frac{1}{N-\ell}\sum\nolimits_{i=\ell}^{N} x_i
$$

## Autocorrelogram (ACF)

:::columns
:::{.column width=60%}
```{r}
ell <- 10
c <- ts_xts(chicken) %>% diff() %>% fortify() %>% mutate(i=1:n(), .before=Index) %>% filter(!is.na(value))

ps <- ggplot(c[ell:length(c$i),]) +
  geom_line(aes(x=i, y=value)) +
  geom_line(data=c[1:ell,], aes(x=i, y=value), color=gray(2/3), linetype=2) +
  coord_cartesian(xlim=range(c$i)) +
  theme(axis.title.x = element_blank(), axis.text.x = element_blank()) +
  labs(y="Value")
pt <- ggplot(lag(c, ell)) +
  geom_line(aes(x=i+ell, y=value)) +
  coord_cartesian(xlim=range(c$i)) +
  labs(x="Index", y="Value")
ggarrange(ps, pt, nrow=2, heights=c(1,1.08))
```
:::
:::{.column width=40%}
To construct the ACF plot of a time series, the axis is shifted by $\ell$ samples, neglecting the first $\ell$ samples in the unshifted series and the last $\ell$ samples in the shifted one

Then the autocorrelation between the two series is computed

The process is repeated for $\ell=0, 1, \dots, n$, with $n$ chosen as a function of the series length, typically at least 50 and in any case no more than half the length of the time series
:::
:::

## Autocorrelogram (ACF)

:::columns
:::column
The **autocorrelation function** $\rho_x(\ell)$, which is a **discrete-valued** function, can be plotted to study the maximum *lag* beyond which the historical series $x$ is no longer autocorrelated

The figure shows the signal of a microphone recording the sound "AAHH". The series is **clearly periodic** every 0.01 s

The autocorrelogram shows **high autocorrelation** up to $\ell=4$. Then the pattern is **periodic**, confirming that the time series is autocorrelated with itself about every 10 lags

:::

:::column
:::panel-tabset
### Time series
```{r}
#| fig-height: 3
s <- ts_data.frame(speech) %>% slice(seq(1,n(),by=1)) %>%
  mutate(time=seq_along(time)/10000)
s %>%
ggplot(aes(x=time, y=value)) + 
  geom_line() +
  labs(x="Time (s)", y="Voltage (mV)")
```

### ACF
```{r}
#| fig-height: 3
with(acf(s$value, lag.max = 100, plot=F), data.frame(lag, acf)) %>%
  ggplot(aes(x=lag, y=acf)) + 
  geom_hline(aes(yintercept=0)) +
  geom_hline(yintercept=c(-1,1)*qnorm((1 + 0.95)/2)/sqrt(1020),
             linetype=2, color="blue") +
  geom_segment(mapping = aes(xend = lag, yend = 0)) +
  labs(x="lag", y=TeX("$\\rho_x$"))
```
:::
:::
:::

:::aside
**ACF** stands for *Auto-Correlation Function*; the blue highlighted band is the 95% confidence interval. Also, in general, $\rho_x(0)=1$ always holds
:::

## Autocorrelogram (ACF)

:::columns
:::column
Consider the same data as the previous time series, but sampled at random instants

Then $y\overset{IID}\sim \mathcal{N}(\mu, \sigma^2)$ and therefore $\rho_y(\ell) = 0~\forall \ell>0$

As expected, the only ACF value outside the confidence interval is $\rho_x(0)$

In this case one also says that the time series is a ***random walk***
:::

:::column
:::panel-tabset
### Time series
```{r}
#| fig-height: 3
s <- s %>% mutate(value = sample(value))
s %>%
ggplot(aes(x=time, y=value)) + 
  geom_line() +
  labs(x="Time", y="Voltage (mV)")
```

### ACF
```{r}
#| fig-height: 3
with(acf(s$value, lag.max = 100, plot=F), data.frame(lag, acf)) %>%
  ggplot(aes(x=lag, y=acf)) + 
  geom_hline(aes(yintercept=0)) +
  geom_hline(yintercept=c(-1,1)*qnorm((1 + 0.95)/2)/sqrt(1020),
             linetype=2, color="blue") +
  geom_segment(mapping = aes(xend = lag, yend = 0)) +
  labs(x="lag", y=TeX("$\\rho_x$"))
```

:::
:::
:::



## Stationary time series
A time series may or may not be stationary. The following are defined:

Time series stationary in the **wide sense**
: It is a time series for which the probabilistic behavior of any collection of values $\left<x_{t_1}, x_{t_2},\dots, x_{t_k}\right>$ is identical to that of the shifted collection $\left<x_{t_1+h}, x_{t_2+h},\dots, x_{t_k+h}\right>$, that is:
$$
\mathrm{Pr}(x_{t_1}\leq c_1, \dots, x_{t_k}\leq c_k) = \mathrm{Pr}(x_{t_1+h}\leq c_1, \dots, x_{t_k+h}\leq c_k).
$$

Time series stationary in the **strict sense**
: It is a time series for which the mean value of the series is constant, independent of time,
  and the autocovariance function $\gamma(s,t)$ depends on $s$ and $t$ only through their difference $|s-t|$ 

For a **wide-sense stationary series** one may assume $\sigma_x(0)=\sigma_x(\ell)=\sigma_x$ and $\bar x_0=\bar x_\ell$ and, therefore, $\rho_x(\ell) = \gamma_x(\ell)/\sigma_x$


## Stabilization

* Time series stationary at least in the wide sense are easier to handle
* It is therefore useful to try to **stabilize** the time series by separating it into a **trend** term $x_t$ plus a **stationary** term $x_s$
* stabilization can be done in two ways:
  - *detrending* by linear regression: $x_t = x_{l,t} + x_{s,t}$, where $x_{l,t} = a + bt$ and, therefore, $x_{s,t}$ is the **residual series** of the linear regression of $x_t$
  - *detrending* by differencing
* The stabilized time series can then be analyzed and then transformed back by the inverse operation:
  - adding the trend term
  - integration, that is cumulative sum
  
## Stabilization (example)
:::{style="text-align: center;"}
```{r fig.width=8}
m <- lm(chicken~time(chicken))
f <- xts(data.frame(
    line = predict(m),
    res = residuals(m)
  ), order.by=as.Date(time(chicken)))
p1 <- ggplot(ts_xts(chicken)) +
  geom_line(aes(x=Index, y=value)) +
  geom_line(aes(x=Index, y=line), data=f, col="blue") +
  labs(title="Price of a Whole Chicken in Georgia (US)", x="Date", y="Price ($)")

p2 <- ggplot(f) +
  geom_hline(yintercept = 0, color="blue") +
  geom_line(aes(x=Index, y=res)) +
  labs(title="Linear De-trend", x="Date", y="Price ($)")

p3 <- ggplot(diff(ts_xts(chicken))) +
  geom_hline(yintercept = 0, color="blue") +
  geom_line(aes(x=Index, y=value)) +
  labs(title="Differenced De-trend", x="Date", y="Price ($)")

p1 / (p2 + p3)
```
:::


## Differencing operators

In general, stabilization by differencing gives better results and is also more practical: if the differenced time series is not stable, the differencing order can be increased until stability is reached

How is a time series differenced?

The following are defined:

* **Backshift operator**: this is the operator $B^n$ such that $B^n x_t := x_{t-n}$
* **Difference operator**: this is the operator $\nabla$ such that $\nabla x_t := x_t - x_{t-1}=(1-B)x_t$. It follows that $\nabla^dx_t=(1-B)^dx_t$, and therefore for example $\nabla^2x_t=(1-B)^2x_t=x_t-2x_{t-1}+x_{t-2}$

So, for example, differencing $\nabla^2x_t$ is the discrete equivalent of the second derivative $\frac{d^2}{dt^2}x(t)$ for the continuous function $x(t)$

## Stabilization and ACF

:::{style="text-align: center;"}
```{r fig.width=8}
p1 <- with(acf(as.numeric(chicken), lag.max = 50, plot=F), data.frame(lag, acf)) %>%
  ggplot(aes(x=lag, y=acf)) + 
  geom_hline(aes(yintercept=0)) +
  geom_hline(yintercept=c(-1,1)*qnorm((1 + 0.95)/2)/sqrt(1020),
             linetype=2, color="blue") +
  geom_segment(mapping = aes(xend = lag, yend = 0)) +
  labs(title="ACF ST Chicken Price", x="lag", y=TeX("$\\rho_x$"))

p2 <- with(acf(f$res, lag.max = 50, plot=F), data.frame(lag, acf)) %>%
  ggplot(aes(x=lag, y=acf)) + 
  geom_hline(aes(yintercept=0)) +
  geom_hline(yintercept=c(-1,1)*qnorm((1 + 0.95)/2)/sqrt(1020),
             linetype=2, color="blue") +
  geom_segment(mapping = aes(xend = lag, yend = 0)) +
  labs(title="ACF linear de-trend", x="lag", y=TeX("$\\rho_x$"))

p3 <- with(acf(diff(chicken), lag.max = 50, plot=F), data.frame(lag=1:51, acf)) %>%
  ggplot(aes(x=lag, y=acf)) + 
  geom_hline(aes(yintercept=0)) +
  geom_hline(yintercept=c(-1,1)*qnorm((1 + 0.95)/2)/sqrt(1020),
             linetype=2, color="blue") +
  geom_segment(mapping = aes(xend = lag, yend = 0)) +
  labs(title="ACF ST Differenced", x="lag", y=TeX("$\\rho_x$"))

p1 / (p2 + p3)
```
:::

## Stabilization and ACF

:::columns

:::column
* The ACF of a non-stabilized series always shows correlation even at high *lags*
* The ACF of the residual series from a linear regression decreases faster, but still keeps correlation even at high *lags*
* The ACF of the differenced series, moreover, damps out very quickly, the *lags* 1, 2, 3, ... should be exponential, after which it shows harmonic oscillations, indicating a **periodicity** in the original time series
:::

:::column
```{r}
p3
```

:::

:::

:::aside
We can therefore say that the historical chicken price series shows a *trend* that **stabilizes at first-order** differencing, has dynamics that show **autocorrelation up to the previous three points**, and a **periodic** pattern with period equal to 12 *lags*.
:::



# Statistical models

Classical regression applied to time series **is often insufficient**. For example, in the case of chicken price, autocorrelation analysis shows a cyclic behavior that classical regression cannot highlight

It is therefore necessary to develop techniques that allow the details of a time series to be modeled, in particular by also taking into account the **autocorrelation** that can characterize time series

## AR models --- definition

An Auto-Regressive (AR) model expresses a given observation $x_t$ at time $t$ as a linear combination of $p$ previous values $x_{t-1}, x_{t-2},\dots,x_{t-p}$. An AR model of order $p$, abbreviated as $\mathrm{AR}(p)$, has the form:
$$
x_t=\phi_1 x_{t-1} + \phi_2 x_{t-2}+\dots+\phi_p x_{t-p} + w_t \label{eq:AR}
$$

* $x_t$ is wide-sense stationary and $w_t\sim\mathcal{N}(0, \sigma^2_w)$
* $\phi_1, \phi_2,\dots,\phi_p$ are constants and $\phi_p\neq0$

If the mean of $x_t$ is not zero, $x_t$ is replaced with $x_t - \mu$ to obtain:
$$
\begin{align}
x_t-\mu&=\phi_1( x_{t-1}-\mu) + \phi_2 (x_{t-2}-\mu)+\dots+\phi_p (x_{t-p}-\mu) + w_t \\
x_t&=\alpha + \phi_1 x_{t-1} + \phi_2 x_{t-2}+\dots+\phi_p x_{t-p} + w_t,~~~\alpha=\mu(1-\phi_1-\phi_2-\dots-\phi_p)
\end{align}
$$

:::aside
The $w_t$ are also called **innovations**, since they are the only **new** contribution of a point with respect to the previous ones.
:::

## AR models --- regression

Recalling the definition of the **backshift operator**, the definition of $w_t$ can be written as:

$$
(1-\phi_1 B - \phi_2 B^2 -\dots-\phi_p B^p)x_t=w_t
$$
o ancora piĂą concisamente come:

$$
\Phi_p(B)x_t=w_t
$$
where $\Phi_p(B):=1-\phi_1B-\phi_2B^2-\dots-\phi_pB^p$ is called the **autoregressive operator**.

Performing the regression of an $\mathrm{AR}(p)$ model on a historical series $x_t$ therefore means fitting the model $\Phi_p(B)\hat{ x_t}=w_t$ by identifying the coefficients of $\Phi_p(B)$ that minimize the mean squared residuals, where the residuals are $\varepsilon_t = x_t - \hat{ x_t} = x_t - \Phi_p^{-1}(B)w_t$, assuming the inverse $\Phi_p^{-1}(B)$ exists.



## MA models --- definition

Alternatively, one can imagine the case in which the generic observation $x_t$ is expressed as a linear combination of the **disturbance** acting on the previous $q$ observations. An MA model of order $q$, abbreviated as $\mathrm{MA}(q)$, is defined as
$$
x_t = w_t + \theta_1 w_{t-1} + \theta_2 w_{t-2} + \dots + \theta_q w_{t-q}
$$

* $w_t\sim\mathcal{N}(0, \sigma_w^2)$
* $\theta_1, \theta_2,\dots,\theta_q$ are constant parameters with $\theta_q\neq0$

## MA models --- regression

Similarly to the $\mathrm{AR}(p)$ case, for $\mathrm{MA}(q)$ it is possible to define the **moving-average operator** of order $q$:
$$
\Theta_q(B)=(1+\theta_1 B+\theta_2 B^2+\dots+\theta_q B^q)
$$ 
such that the equation for $x_t$ can be written as:

$$
x_t=\Theta_q(B)w_t
$$

As above, regressing an $\mathrm{MA}(q)$ model onto a historical series $x_t$ means identifying the terms of $\Theta_q(B)$ that minimize the mean squared residuals, defined as $\varepsilon_t = x_t - \hat{x_t} = x_t - \Theta_q(B)$, note that this time there is no inverse.


## MA models and ACF
```{r}
set.seed(0)
tsma <- arima.sim(model=list(ma=c(-0.5, 0.7, -0.75)), n=200)
```

:::columns

:::column
An $MA(q)$ time series has a memory extending up to *lag* $q$, in the sense that innovations at distances greater than $q$ have no effect on the last observation

Therefore, given an $MA(q)$ time series, one can often identify the order from its ACF by counting the peaks after the one at *lag* 0:

* three peaks outside the confidence band mean an $MA(3)$ model
* the signs of the peaks correspond to the signs of the coefficients


:::

:::column


```{r}
autoplot(tsma, xlab="#", ylab="Value") /
(autoplot(acf(tsma, plot=F)) + 
  geom_hline(yintercept=0))
```
:::

:::


## MA models and ACF
```{r}
set.seed(0)
tsar <- arima.sim(model=list(ar=c(0.5, 0.3)), n=200)
```

:::columns

:::column
If instead the time series is of type $AR(p)$, each observation depends on the innovation and on all previous observations, recursively

In this case the autocorrelogram will show an exponential decay possibly followed by harmonic oscillations

:::

:::column


```{r}
autoplot(tsar, xlab="#", ylab="Value") /
(autoplot(acf(tsar, plot=F)) + 
  geom_hline(yintercept=0))
```
:::

:::


## AR models and PACF

For models of type $AR(p)$ it is still possible to identify the order through the **partial autocorrelation function**, or PACF, defined as follows:

$$
\begin{align}
\mathrm{PACF}_1 &= \mathrm{ACF}(z_{t+1}, z_t) \\
\mathrm{PACF}_k &= \mathrm{ACF}(z_{t+k} - \hat{z_{t+k}}, z_t - \hat{z_t}),~~~k\geq2 \\
\end{align}
$$
where $\hat{z_{t+k}}$ and $\hat{z_t}$ are linear combinations of $\{z_{t+1}, z_{t+2},\dots,z_{t+k-1}\}$ that minimize the mean squared error of $z_{t+k}$ and $z_t$, respectively



## AR models and PACF

:::columns

:::column
In general, therefore, if the ACF shows an infinite memory, AR model, and the PACF shows few peaks, the number of peaks is the order of the AR model

**Warning**: peaks after the first *cut-off* are not considered, that is, the *lag* at which the autocorrelation falls below the confidence limit for the first time
:::

:::column
```{r}
forecast::ggtsdisplay(tsar, main="AR Series")
```
:::

:::


:::aside
**Warning**: in these ACF and PACF plots the *lag* starts from 1.
:::

## MA models and PACF

:::columns

:::column
Comparing both ACF and PACF

* pure noise: ACF and PACF are zero for *lag* greater than 0
* $AR(p)$: ACF decreases slowly, PACF is non-zero for *lags* less than or equal to $p$, zero otherwise
* $MA(q)$: ACF shows $q$ peaks; if $PACF(1) > 0$, PACF oscillates to 0, otherwise it decays geometrically to 0
:::

:::column
```{r}
forecast::ggtsdisplay(tsma, main="MA Series")
```
:::
:::

:::aside
Often, in practice, these indications are hard to observe and it is therefore difficult to identify $p,q$ in a certain and unambiguous way
:::


## ARMA models

The obvious extension comes from combining the $\mathrm{AR}(p)$ and $\mathrm{MA}(q)$ models

An ARMA model of order $p,q$, abbreviated as $\mathrm{ARMA}(p,q)$, is defined as:

$$
x_t=\phi_1 x_{t-1} + \phi_2 x_{t-2}+\dots+\phi_p x_{t-p} + w_t + \theta_1 w_{t-1} + \theta_2 w_{t-2} + \dots + \theta_q w_{t-q}
$$
con $\phi_p$ e $\theta_q$ non nulli, ovvero, piĂą brevemente:
$$
\Phi_p(B)x_t=\Theta_q(B)w_t
$$

## MA models and PACF

:::columns

:::column
Comparing both ACF and PACF

* pure noise: ACF and PACF are zero for *lag* greater than 0
* $AR(p)$: ACF decreases slowly, PACF is non-zero for *lags* less than or equal to $p$, zero otherwise
* $MA(q)$: ACF shows $q$ peaks; if $PACF(1) > 0$, PACF oscillates to 0, otherwise it decays geometrically to 0
* $ARMA(p,q)$: ACF shows $q$ peaks; PACF decays geometrically to 0 only after *lag* $p$
:::

:::column
```{r}
forecast::ggtsdisplay(tsma, main="MA Series")
```
:::
:::

:::aside
Often, in practice, these indications are hard to observe and it is therefore difficult to identify $p,q$ in a certain and unambiguous way
:::


## ARIMA models
The $\mathrm{ARMA}(p,q)$ processes are suitable only for describing **stationary** series in the wide sense

We have seen, however, that a non-stationary process can be made stationary by differencing of a suitable degree $d$

A process $x_t$ is called $\mathrm{ARIMA}(p,d,q)$ when $\nabla^d x_t = (1-B)^d x_t$ is an $\mathrm{ARMA}(p,q)$ process

In general, when $E(\nabla^dx_t)=0$, the $\mathrm{ARIMA}(p,d,q)$ process can be written as:
$$
\Phi_p(B)\nabla^dx_t=\Theta_q(B)w_t
$$
If instead the expected value $E(\nabla^dx_t)=\mu$, then:
$$
\Phi_p(B)\nabla^dx_t=\delta + \Theta_q(B)w_t,~~~\delta=\mu(1-\phi_1-\dots-\phi_p)
$$

## Stability and uniqueness of models

Consider an $\mathrm{AR}(1)$ model: we can expand the recursive formula as:
$$
\begin{align}
x_t =& \phi x_{t-1}+w_t = \phi(\phi x_{t-2}+w_{t-1}) + w_t \\
=& \sum_{j=0}^{+\infty}\phi^j w_{t-j}
\end{align}
$$
It is therefore clear that the time series $x_t$ **is stable only if** $|\phi|<1$

In the general case of the $\mathrm{AR}(p)$ model, the stability condition is:
$$
\left|\sum_{i=1}^p \phi_i\right| < 1 \label{eq:ARstab}
$$
An unstable AR model is also called **anti-causal**, because it can be shown that to be stabilized it requires knowledge of future observations

## Stability and uniqueness of models

For an $\mathrm{MA}(1)$ model, instead, consider two models:
$$
\begin{align}
x_t &= w_t + \theta w_{t-1}, ~w_t\sim\mathcal{N}(0, 1) \\
y_t &= \nu_t + 1/\theta \nu_{t-1}, ~\nu_t\sim\mathcal{N}(0, \theta^2) 
\end{align}
$$
It is clear that $x_t$ and $\nu_t$ have the same ACF and are indistinguishable, since we do not know the innovations $w_t$ and $\nu_t$ but only the two series. It can be shown that the result is general, that is, it does not depend on the order $q$. 

It is therefore necessary to choose one of the two alternative forms. To determine which one, we rewrite the series as $w_t=-\theta x_{t-1}+x_t$, which has the same form as $\mathrm{AR}(1)$. We can then choose the alternative that satisfies the same *stability* criterion:
$$
\left|\sum_{i=1}^q \theta_i\right| < 1
$$
This alternative is called **invertible**


## Stability and uniqueness of models

In general, it can be shown that the criteria seen above correspond to imposing the requirement that the complex roots of the polynomials 
$$
\begin{align}
\Phi_p(z)&=1-\phi_1 z-\phi_2 z^2 - \dots -\phi_p z^p \\
\Theta_q(z)&=1+\theta_1 z+\theta_2 z^2 + \dots +\theta_q z^q
\end{align}
$$ 
are **all strictly outside the unit circle** in the complex plane

For example: $\mathrm{ARMA}(2,2)$ with $\Phi_2(z)= 1 + 0.9z + 0.1z^2$ and $\theta_q(z) = 1 + 2z + 15z^2$:
```{r echo=T}
abs(polyroot(c(1, -0.9, -0.1)))
abs(polyroot(c(1, 2, 15)))
```

That is, the AR term is not causal and the MA term is not invertible, but $\theta_q(z) = 1 + 1/2z + 1/15z^2$ would be


## Redundancy

Consider the process $x_t = w_t$, with $w_t\sim\mathcal{N}(0, 1)$: this is obviously pure random noise

Multiply both sides by $1-0.5B$ to obtain:
$$
x_t = 0.5x_{t-1}-0.5w_{t-1} + w_t
$$
which **looks like** an $\mathrm{ARMA}(1,1)$ process but is obviously still the same random noise. **How do we distinguish these cases?**

* factorize the polynomials $\Phi_p$ and $\Theta_q$
* remove the **common factors**
* recombine the factors to obtain the **non-redundant model**

In R one can still use `polyroot()`

## Redundancy

For example, consider the process $x_t=0.4x_{t-1}+0.45x_{t-2}+w_t + w_{t-1}+0.25w_{t-2}$, which using the operator $B$ becomes:
$$
(1-0.4B-0.45B^2)x_t = (1+B+0.25B^2)w_t.
$$

In this form the process looks like $\mathrm{ARMA}(2,2)$, however we can factor the two polynomials using `polyroot()` to compute the roots:
```{r echo=T}
# per Phi:
-1/polyroot(c(1, -0.4, -0.45))
# per Theta:
-1/polyroot(c(1, 1, 0.25))
```
:::aside
**Note**: if $z_i,~ i=1,\dots,n$ are the roots of the polynomial $p(z)$ of degree $n$, then the polynomial can be factorized into $(1-1/z_1 z)\cdot(1-1/z_2 z)\dots(1-1/z_n z)$.
:::

## Redundancy

As can be seen, one can write 
$$
\begin{align}
\Phi_p(B)&=(1-0.9B)(1+0.5B) \\
\Theta_q(B)&=(1+0.5B)^2
\end{align}
$$

Removing the common factor $(1+0.5B)$ we obtain the model 
$$
x_t=0.9x_{t-1}+0.5w_{t-1}+w_t
$$
which is an $\mathrm{ARMA}(1,1)$.

Therefore, in addition to the criteria of **causality** and **invertibility**, one must add the criterion of **non-redundancy** of the parameters, which is checked by removing every common factor from the factorization of the polynomials $\Phi_p(B)$ and $\Theta_q(B)$.

## (S)ARIMA regression

From what has been said above, a process, or time series, $x_t$ can be regressed by means of an ARIMA model by identifying its parameters through **minimization of squared deviations**

However, before regression it is necessary to define suitable indices $p, d,q$, such that there is **neither underfitting nor overfitting**

In some cases, moreover, series are **periodic**: besides a possible **trend** they are also subject to cyclic oscillating patterns. In these cases:

* the time series is made stationary by differencing
* a low-frequency contribution, called **seasonal**, is separated from a high-frequency contribution
* the seasonal and non-seasonal contributions are regressed separately as two distinct and superposed ARIMA processes: this regression is called *Seasonal ARIMA*, or **SARIMA**

## (S)ARIMA regression

Moreover, we have seen that ARIMA models are based on the assumption of time series stationary in the wide sense

Therefore, it is necessary that both the **mean value and the variance are constant over time**

* The mean is stabilized by differencing
* The variance can be stabilized through **Box-Cox transformations**

Therefore, in the most general case, the model is $\mathrm{SARIMA}(p, d, q, p_s, d_s, q_s, \lambda)$

**Before performing a regression it is therefore necessary to define the values of the seven parameters, avoiding overfitting and underfitting**

* $d, d_s$ can be identified by trial, gradually increasing the values, first $d$ and then $d_s$, until the ACF is satisfactory
* the other terms can be identified by checking the goodness of the regression over a grid of combinations and evaluating a suitable **figure of merit**

## Akaike Information Criterion

The most used figure of merit in SARIMA regression is the **Akaike Information Criterion**, or AIC

In general, the quality of a regression with $k$ parameters is measured by the sum of squared residuals, $SS_E(k)$, normalized by the sample size $n$:
$$
\hat{\sigma_k^2} = \frac{SS_E(k)}{n}
$$
This indicator is called the ***Maximum Likelihood Estimator***, MLE. The smaller this value is, the better the model fits the data

However, increasing $k$ decreases the MLE, at the risk of overfitting. For this reason, Akaike proposed penalizing the MLE with the number of parameters:
$$
\mathrm{AIC}=\log(\hat{\sigma_k^2})+\frac{n+2k}{n}
$$
The AIC must obviously be **minimized**

## Other information criteria

Besides the AIC there are also the corrected AIC and the *Bayesian Information Criterion*, or BIC:

* Corrected AIC: $\mathrm{AIC}_c = \log(\hat{\sigma_k^2})+\frac{n+k}{n-k-2}$
* BIC: $\mathrm{BIC} = \log(\hat{\sigma_k^2})+\frac{k\log(n)}{n}$

The BIC **penalizes model size more strongly**, and is therefore preferred for very large samples, thousands of observations, for which AIC and AICc would tend to favor unnecessarily complex models, too many parameters, overfitting

Like the AIC, AICc and BIC must also be **minimized**

## Regression

The R library `forecast` provides the function `auto.arima()`, which evaluates the indicators over a grid of combinations of the seven parameters and returns the best regression:

:::columns
:::{.column style="font-size: 70%;"}
```{r echo=T}
forecast::auto.arima(AirPassengers, lambda="auto", trace=T)
```
:::

:::column

:::panel-tabset
### Time series
```{r}
AirPassengers %>% autoplot() +
  labs(x="Time", y="Number of passengers", title="US Airline Passengers")
```
### Stabilized
```{r}
diff(AirPassengers^-0.295) %>% autoplot() +
  labs(x="Time", y="Number of passengers", title="US Airline Passengers")
```
:::
:::
:::

## Forecasting

:::columns

:::column
Once the most suitable model has been selected, the one with **minimum AIC**, one can proceed with regression and prediction

Prediction can be obtained with the `forecast()` function

This function also reports the 95% confidence bands
:::

:::column
:::panel-tabset
# Forecast
```{r}
fit <- forecast::auto.arima(AirPassengers, lambda="auto")
autoplot(forecast::forecast(fit,h=20)) +
  theme(axis.text.x = element_text(angle = 45,
        hjust = 1)) +
  labs(title="ARIMA(0,1,1)(0,1,1)[12] Forecast", x="Time", y="Number of passengers")
```

# Detail
```{r warning=FALSE}
fit <- forecast::auto.arima(AirPassengers, lambda="auto")
forecast::forecast(fit,h=20) %>% 
  ggplot(aes(x=Index, y=Data)) + 
  geom_line() + 
  geom_ribbon(aes(ymin=`Lo 95`, ymax=`Hi 95`), alpha=1/3) +
  geom_ribbon(aes(ymin=`Lo 80`, ymax=`Hi 80`), alpha=1/3) +
  geom_line(aes(y=`Point Forecast`), color="blue") +
  scale_x_date(limits=c(ymd("1960/1/1"), NA), date_minor_breaks = "1 month")+
  theme(axis.text.x = element_text(angle = 45,
        hjust = 1)) +
  labs(title="ARIMA(0,1,1)(0,1,1)[12] Detail", x="Time", y="Number of passengers")
```
:::

:::

:::
