---
title: "Feature-based analysis of time series"
subtitle: "Multi-agent system for smart machining, part II"
author: "Paolo Bosetti"
format: revealjs
---


```{r setup}
options(width = 60)
set.seed(0)
library(latex2exp)
library(tidyverse)
library(ggfortify)
library(patchwork)
library(e1071)
theme_set(theme_gray() + theme(legend.position = "bottom"))

n <- 240
t <- 1:n
fault_idx <- 151:210

x1 <- cumsum(rnorm(n, 0, 0.06)) + 0.7 * sin(2 * pi * t / 24)
x2 <- 0.8 * x1 + rnorm(n, 0, 0.18)
x1[fault_idx] <- x1[fault_idx] + 1.0
x2[fault_idx] <- -0.2 * x1[fault_idx] + rnorm(length(fault_idx), 0, 0.22) + 0.7

ts_df <- tibble(
  t = t,
  x1 = x1,
  x2 = x2,
  state = factor(if_else(t %in% fault_idx, "Fault", "Normal"))
)

pca_fit <- prcomp(ts_df %>% select(x1, x2), center = TRUE, scale. = TRUE)
scores <- as_tibble(pca_fit$x) %>%
  mutate(t = ts_df$t, state = ts_df$state)

lambda <- pca_fit$sdev^2
center_vec <- pca_fit$center
scale_vec <- pca_fit$scale
x_scaled <- scale(ts_df %>% select(x1, x2), center = center_vec, scale = scale_vec)
x_hat <- tcrossprod(pca_fit$x[, 1, drop = FALSE], pca_fit$rotation[, 1, drop = FALSE])
residuals_mat <- x_scaled - x_hat
spe <- rowSums(residuals_mat^2)
t2 <- rowSums((pca_fit$x[, 1:2, drop = FALSE]^2) / matrix(rep(lambda[1:2], each = n), ncol = 2))
alpha_t2 <- 0.05
a_t2 <- 2
n_ref <- sum(ts_df$state == "Normal")
t2_limit <- a_t2 * (n_ref - 1) / (n_ref - a_t2) *
  qf(1 - alpha_t2, df1 = a_t2, df2 = n_ref - a_t2)

ts_pca <- ts_df %>%
  mutate(PC1 = pca_fit$x[, 1],
         PC2 = pca_fit$x[, 2],
         SPE = spe,
         T2 = t2,
         T2_limit = t2_limit)

svm_fit <- svm(state ~ x1 + x2, data = ts_df,
               kernel = "radial", cost = 5, gamma = 1,
               scale = TRUE)
ts_df$pred <- predict(svm_fit, ts_df)
acc <- mean(ts_df$pred == ts_df$state)

grid <- expand.grid(
  x1 = seq(min(ts_df$x1) - 0.3, max(ts_df$x1) + 0.3, length.out = 160),
  x2 = seq(min(ts_df$x2) - 0.3, max(ts_df$x2) + 0.3, length.out = 160)
)
grid$pred <- predict(svm_fit, grid)
```

# Feature-based time-series analysis

In feature-based analysis, a multivariate time series in $p$ dimensions is transformed into a sequence of vectors
$$
\mathbf{x}_t \in \mathbb{R}^p,\quad t = 1,\dots,N
$$
where each component is either a direct measurement or an extracted descriptor such as RMS, variance, spectral energy, kurtosis, or temperature gradient

The objective is to **project, compress, classify**, or separate process states in the feature space rather than directly on the raw waveform

Two standard techniques are:

* **PCA**, mainly for dimensionality reduction, monitoring, and latent-structure analysis
* **SVM**, mainly for supervised classification and margin-based separation

# PCA theory

## PCA: data matrix and centering
::: {.columns}
::: {.column}
Let the **feature matrix** of a MVTS be
$$
\mathbf{X} =
\begin{bmatrix}
\mathbf{x}_1^T \\
\mathbf{x}_2^T \\
\vdots \\
\mathbf{x}_N^T
\end{bmatrix}
\in \mathbb{R}^{N \times p}
$$
where each row is a **feature vector** extracted at one time instant or from one time window statistics
:::
::: {.column}
PCA starts from the centered matrix $\mathbf{X}_c$:
$$
\mathbf{X}_c = \mathbf{X} - \boldsymbol{\mu}^T
$$
with sample means vector:
$$
\boldsymbol{\mu} = \frac{1}{N}\sum_{t=1}^{N}\mathbf{x}_t
$$

If variables have different physical units, one usually **standardizes** them before PCA
:::
:::
<!-- end columns -->

## PCA: covariance matrix and eigenproblem

::: {.columns}
::: {.column}
The sample covariance matrix ($p\times p$)is:
$$
\mathbf{S} = \frac{1}{N-1}\mathbf{X}_c^T \mathbf{X}_c
$$

PCA solves the eigenvalue problem:
$$
\mathbf{S}\mathbf{p}_i = \lambda_i \mathbf{p}_i,
\quad i = 1,\dots,p
$$

:::
::: {.column}

where:

* $\lambda_i$ is the variance explained by the $i$-th principal component
* $\mathbf{p}_i$ is the corresponding eigenvector, or **loading vector**

The indexes $i$ are sorted so that:
$$
\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_p \ge 0
$$

:::
:::
<!-- end columns -->

::: {.callout-note}
The eigenvectors $\mathbf{p}_i$ are orthogonal and form a new coordinate system in feature space, where the first axis captures the largest variance, the second axis captures the largest remaining variance, and so on.
:::


## PCA: scores and low-dimensional representation

::: {.columns}
::: {.column}
The projection of the data onto the principal axes gives the **scores matrix** ($N\times p$):
$$
\mathbf{T} = \mathbf{X}_c \mathbf{P}
$$
where:
$$
\mathbf{P} = [\mathbf{p}_1,\mathbf{p}_2,\dots,\mathbf{p}_p]
$$

:::
::: {.column}
Using only the first $a < p$ components yields the reduced model
$$
\mathbf{X}_c \approx \mathbf{T}_a \mathbf{P}_a^T
$$
where:
$$
\mathbf{T}_a = \mathbf{X}_c \mathbf{P}_a
$$
:::
:::
<!-- end columns -->

::: {.callout-note}
For a given subset of principal components $a$, the **explained variance ratio** is:
$\eta_a = \frac{\sum_{i=1}^{a}\lambda_i}{\sum_{i=1}^{p}\lambda_i}$. This can be used to select the number of components to retain, for example by setting a threshold such as $\eta_a \ge 0.95$.
:::

:::aside
**Note**: We can use transpose rather than inverse because $\mathbf{P}$ is an orthonormal matrix, so $\mathbf{P}^{-1} = \mathbf{P}^T$.
:::

## PCA: reconstruction error and monitoring statistics

::: {.columns}
::: {.column}
For one centered observation vector $\mathbf{x}_{c,t}$, PCA first computes the score vector
$$
\mathbf{t}_t = \mathbf{P}_a^T \mathbf{x}_{c,t}
$$
and then reconstructs it in feature space as
$$
\hat{\mathbf{x}}_{c,t} = \mathbf{P}_a \mathbf{t}_t
$$

Therefore the residual decomposition is:
$$
\mathbf{x}_{c,t} = \hat{\mathbf{x}}_{c,t} + \mathbf{\varepsilon}_t
$$
where $\hat{\mathbf{x}}_{c,t}$ lies in the retained **latent subspace** and $\mathbf{\varepsilon}_t$ is the residual vector
:::
::: {.column}

Two standard **monitoring statistics** are:

* Hotelling's $T^2$:
$$
T_t^2 = \sum_{i=1}^{a}\frac{t_{ti}^2}{\lambda_i}
$$

* Squared Prediction Error, SPE or $Q$ statistic:
$$
Q_t = \|\mathbf{\varepsilon}_t\|_2^2
    = \sum_{j=1}^{p}\varepsilon_{tj}^2
$$
:::
:::
<!-- end columns -->

::: {.callout-note}
$T^2$ measures abnormality inside the retained subspace, while $Q_t$ measures lack of fit outside it
:::


## PCA: why it is useful for time series

PCA is applied to time-series analysis after building features from samples or windows

Typical uses are:

* reducing a multivariate signal to a few latent variables
* detecting operating regimes and structural changes
* decorrelating strongly coupled measurements
* identifying dominant directions of variability

For dynamic systems, PCA alone is not a time model. It is a **feature-space model**, and temporal structure enters through the chosen features or through lag embedding

# SVM theory

## SVM: supervised binary classification

Assume labeled feature vectors
$$
(\mathbf{x}_i, y_i),\quad y_i \in \{-1,+1\},\quad i=1,\dots,N
$$

The goal is to learn a decision function
$$
f(\mathbf{x}) = \mathrm{sign}(g(\mathbf{x}))
$$
that separates the two classes with maximum robustness

For the linear case,
$$
g(\mathbf{x}) = \mathbf{w}^T \mathbf{x} + b
$$
and the separating boundary is the hyperplane
$$
\mathbf{w}^T \mathbf{x} + b = 0
$$

## SVM: maximum-margin principle

If the classes are linearly separable, the hard-margin SVM solves:
$$
\min_{\mathbf{w},b}\frac{1}{2}\|\mathbf{w}\|_2^2
$$
subject to
$$
y_i(\mathbf{w}^T\mathbf{x}_i+b)\ge1,\quad i=1,\dots,N
$$

The geometric margin is
$$
\gamma = \frac{2}{\|\mathbf{w}\|_2}
$$

Therefore, minimizing $\|\mathbf{w}\|_2^2$ is equivalent to maximizing the separation margin

## SVM: soft margin and hinge loss

Real process data are noisy, overlapping, and rarely perfectly separable

The soft-margin formulation introduces slack variables $\xi_i \ge 0$:
$$
\min_{\mathbf{w},b,\boldsymbol{\xi}}
\frac{1}{2}\|\mathbf{w}\|_2^2 + C\sum_{i=1}^{N}\xi_i
$$
subject to
$$
y_i(\mathbf{w}^T\mathbf{x}_i+b)\ge1-\xi_i
$$

Equivalent empirical-risk view:
$$
\min_{\mathbf{w},b}
\frac{1}{2}\|\mathbf{w}\|_2^2 +
C\sum_{i=1}^{N}\max(0,1-y_i(\mathbf{w}^T\mathbf{x}_i+b))
$$

The parameter $C$ balances margin width against classification violations

## SVM: dual problem and kernel trick

::: {.columns}
::: {.column}

The dual formulation is
$$
\max_{\boldsymbol{\alpha}}
\sum_{i=1}^{N}\alpha_i -
\frac{1}{2}\sum_{i=1}^{N}\sum_{j=1}^{N}
\alpha_i\alpha_j y_i y_j K(\mathbf{x}_i,\mathbf{x}_j)
$$
with constraints
$$
\alpha_i \ge 0,\qquad
\sum_{i=1}^{N}\alpha_i y_i = 0
$$

:::
::: {.column}

The kernel function
$$
K(\mathbf{x}_i,\mathbf{x}_j)=\phi(\mathbf{x}_i)^T\phi(\mathbf{x}_j)
$$
allows nonlinear separation in the original space through linear separation in a transformed feature space

Common kernels:

* linear: $K(\mathbf{x},\mathbf{z})=\mathbf{x}^T\mathbf{z}$
* polynomial
* radial basis function, RBF:
$$
K(\mathbf{x},\mathbf{z})=\exp(-\gamma\|\mathbf{x}-\mathbf{z}\|_2^2)
$$

:::
:::
<!-- end columns -->

## SVM: decision function and time-series use

After training, the decision function is
$$
g(\mathbf{x}) = \sum_{i \in SV}\alpha_i y_i K(\mathbf{x}_i,\mathbf{x}) + b
$$
where only the **support vectors** contribute

For time-series analysis, SVM is applied to features extracted from samples or windows, for example:

* direct multivariate measurements $(x_{1,t},x_{2,t},\dots)$
* lagged vectors $(x_t,x_{t-1},\dots,x_{t-L})$
* statistical and spectral features over a window

SVM is therefore a classification rule in feature space, not a stochastic time model

# PCA example in R

## Synthetic bivariate time series

:::columns
:::column
We construct a bivariate time series with two process variables:

* a **normal** regime with strong positive correlation
* a **fault** regime with mean shift and changed covariance structure

This is enough to illustrate how PCA detects changes in latent structure
:::

:::column
```{r}
p1 <- ggplot(ts_df, aes(x = t, y = x1)) +
  geom_line(aes(color="Normal")) +
  geom_line(aes(color="Fault"), data = filter(ts_df, state == "Fault"), linewidth = 1) +
  labs(x = "Time index", y = "x1")

p2 <- ggplot(ts_df, aes(x = t, y = x2)) +
  geom_line(aes(color="Normal")) +
  geom_line(aes(color="Fault"), data = filter(ts_df, state == "Fault"), linewidth = 1) +
  labs(x = "Time index", y = "x2")

p1 / p2
```
:::
:::

## PCA on the bivariate series

:::columns
:::column
We apply PCA to the standardized two-variable feature matrix

Since $p=2$, the model is simple enough to visualize directly:

* **loadings** describe the principal directions
* **scores** show how observations evolve in the latent space
* **explained variance** quantifies dimensional compression
:::

:::column
```{r}
summary(pca_fit)
round(pca_fit$rotation, 3)
```

```{r}
autoplot(pca_fit, data = ts_df, colour = "state",
         loadings = TRUE, loadings.label = TRUE,
         loadings.label.size = 3) +
  labs(title = "PCA score plot with loadings") +
  coord_fixed()
```
:::
:::

## PCA monitoring statistics

:::columns
:::column
For each time instant we compute:

* score coordinates PC1 and PC2
* SPE, squared prediction error
* Hotelling's $T^2$

For $T^2$ we can define a control limit associated with a false-alarm probability $\alpha$:
$$
T^2_{\alpha} =
\frac{a(n-1)}{n-a}F_{a,n-a;1-\alpha}
$$
where $a$ is the number of retained components and $n$ is the reference sample size

:::

:::column
```{r}
p1 <- ggplot(ts_pca, aes(x = t)) +
  geom_line(aes(y = PC1, color = "Normal")) +
  geom_line(aes(y = PC2, color = "Normal"), alpha = 0.5) +
  geom_line(aes(y = PC1, color = "Fault"), data = filter(ts_pca, state == "Fault"), linewidth = 1) +
  geom_line(aes(y = PC2, color = "Fault"), data = filter(ts_pca, state == "Fault"), linewidth = 1, alpha = 0.5) +
  labs(x = "Time index", y = "PC1/PC2") + 
  theme(legend.position = "none")

p2 <- ggplot(ts_pca, aes(x = t, y = SPE)) +
  geom_line(aes(color = "Normal")) +
  geom_line(aes(color = "Fault"), data = filter(ts_pca, state == "Fault"), linewidth = 1) +
  labs(x = "Time index", y = "SPE") +
  theme(legend.position = "none")

p3 <- ggplot(ts_pca, aes(x = t, y = T2)) +
  geom_line(aes(color = "Normal")) +
  geom_line(aes(color = "Fault"), data = filter(ts_pca, state == "Fault"), linewidth = 1) +
  geom_hline(yintercept = t2_limit, color = "blue", linetype = 2) +
  labs(x = "Time index", y = TeX("$T^2$"))

p1 / p2 / p3
```
:::
:::

:::aside
In this synthetic example, PCA is mainly used as an exploratory and monitoring tool: it reveals the regime change without using class labels during training.
:::

# SVM example in R

## SVM on the same bivariate time series

:::columns
:::column
Now we use the same observations, but with labels `Normal` and `Fault`

The feature vector is simply
$$
\mathbf{x}_t = [x_{1,t},x_{2,t}]^T
$$

An RBF-kernel SVM is trained to separate the two states
:::

:::column
```{r}
svm_fit
acc
```

```{r}
ggplot() +
  geom_tile(data = grid,
            aes(x = x1, y = x2, fill = pred),
            alpha = 0.25) +
  geom_point(data = ts_df,
             aes(x = x1, y = x2, color = state),
             size = 1.8) +
  labs(x = "x1", y = "x2",
       title = "SVM decision regions on the feature space")
```
:::
:::

## SVM predicted labels over time

:::columns
:::column
The decision boundary is learned in feature space, but the output can be mapped back to time

This makes it possible to compare:

* true process state
* SVM predicted state
* classification errors and transition regions
:::

:::column
```{r}
ggplot(ts_df, aes(x = t)) +
  geom_line(aes(y = as.numeric(state), color = "True"), linewidth = 1) +
  geom_line(aes(y = as.numeric(pred), color = "Predicted"),
            linewidth = 0.9, linetype = 2) +
  scale_y_continuous(breaks = c(1, 2),
                     labels = levels(ts_df$state)) +
  labs(x = "Time index", y = "Class",
       color = "", title = "True and predicted labels")
```
:::
:::

## Python example: same synthetic dataset

The same bivariate time series can be generated in Python with `numpy` and stored in a `pandas` data frame

```python
import numpy as np
import pandas as pd

np.random.seed(0)

n = 240
t = np.arange(1, n + 1)
fault_idx = np.arange(150, 210)

x1 = np.cumsum(np.random.normal(0, 0.06, n)) \
     + 0.7 * np.sin(2 * np.pi * t / 24)
x2 = 0.8 * x1 + np.random.normal(0, 0.18, n)

x1[fault_idx] = x1[fault_idx] + 1.0
x2[fault_idx] = -0.2 * x1[fault_idx] \
                + np.random.normal(0, 0.22, len(fault_idx)) + 0.7

state = np.where(np.isin(t, fault_idx + 1), "Fault", "Normal")

df = pd.DataFrame({
    "t": t,
    "x1": x1,
    "x2": x2,
    "state": state
})
```

# PCA example in Python

## PCA on the bivariate series in Python

:::columns
:::column
The PCA workflow in Python is analogous to the R one:

* standardize the variables
* fit a PCA model
* inspect explained variance and loadings
* project the observations to obtain principal scores
:::

:::column
```python
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA

X = df[["x1", "x2"]].to_numpy()

scaler = StandardScaler()
X_scaled = scaler.fit_transform(X)

pca = PCA(n_components=2)
scores = pca.fit_transform(X_scaled)

df["PC1"] = scores[:, 0]
df["PC2"] = scores[:, 1]

print(pca.explained_variance_ratio_)
print(pca.components_)
```
:::
:::

## PCA monitoring statistics in Python

:::columns
:::column
As in R, we can reconstruct the retained subspace and compute:

* the score trajectories
* Hotelling's $T^2$
* the residual SPE statistic

This provides a direct parallel with the PCA monitoring slide shown earlier
:::

:::column
```python
import numpy as np

lam = pca.explained_variance_
X_hat = scores @ pca.components_
E = X_scaled - X_hat

df["SPE"] = np.sum(E**2, axis=1)
df["T2"] = np.sum((scores**2) / lam, axis=1)
```

```python
import matplotlib.pyplot as plt

fig, ax = plt.subplots(2, 1, figsize=(8, 5), sharex=True)
ax[0].plot(df["t"], df["PC1"])
ax[0].set_ylabel("PC1")
ax[1].plot(df["t"], df["SPE"])
ax[1].set_ylabel("SPE")
ax[1].set_xlabel("Time index")
```
:::
:::

# SVM example in Python

## SVM on the same bivariate time series in Python

:::columns
:::column
Using `scikit-learn`, the same classification problem is solved by fitting an RBF-kernel SVM to the two features

The workflow is:

* build `X` and class labels `y`
* fit `SVC`
* predict the class of each observation
* evaluate accuracy and inspect the decision regions
:::

:::column
```python
from sklearn.svm import SVC
from sklearn.metrics import accuracy_score

X = df[["x1", "x2"]].to_numpy()
y = df["state"].to_numpy()

svc = SVC(kernel="rbf", C=5.0, gamma=1.0)
svc.fit(X, y)

df["pred"] = svc.predict(X)
acc = accuracy_score(y, df["pred"])
print(acc)
```
:::
:::

## SVM decision regions and predicted labels in Python

:::columns
:::column
The decision map in feature space can be visualized by evaluating the classifier on a regular grid

The predicted labels can then be plotted back against time exactly as done in the R example
:::

:::column
```python
x1g = np.linspace(df["x1"].min() - 0.3, df["x1"].max() + 0.3, 160)
x2g = np.linspace(df["x2"].min() - 0.3, df["x2"].max() + 0.3, 160)
xx1, xx2 = np.meshgrid(x1g, x2g)
Xg = np.column_stack([xx1.ravel(), xx2.ravel()])
zg = svc.predict(Xg).reshape(xx1.shape)
```

```python
fig, ax = plt.subplots(1, 2, figsize=(10, 4))
ax[0].contourf(xx1, xx2, (zg == "Fault").astype(float), alpha=0.25)
ax[0].scatter(df["x1"], df["x2"], c=(df["state"] == "Fault"))
ax[0].set_xlabel("x1")
ax[0].set_ylabel("x2")

ax[1].plot(df["t"], df["state"].eq("Fault").astype(int), label="True")
ax[1].plot(df["t"], df["pred"].eq("Fault").astype(int), "--",
           label="Predicted")
ax[1].set_xlabel("Time index")
ax[1].legend()
```
:::
:::

## PCA and SVM: complementary roles

PCA and SVM serve different purposes:

* PCA is **unsupervised** and looks for directions of maximum variance
* SVM is **supervised** and looks for a boundary that best separates known classes

In practice they are often combined:

* PCA for denoising or dimensionality reduction
* SVM for classification on the retained principal scores

For high-dimensional time-series features this PCA plus SVM pipeline is often more stable than direct classification on raw variables alone
