---
title: "01 - PCA Example on Manufacturing Signals"
format: pdf
execute:
  echo: true
  warning: false
  message: false
---

## Objective

Use PCA to reduce multivariate sensor features before state classification.

```{r}
library(tidyverse)
library(tidymodels)
```

## Simulated manufacturing signals

```{r}
set.seed(42)

n <- 1200
states <- c("idle", "stable", "anomaly")

dat <- tibble(
  state = sample(states, n, replace = TRUE, prob = c(0.35, 0.50, 0.15)),
  power = rnorm(n, if_else(state == "idle", 20, if_else(state == "stable", 55, 75)), 6),
  vibration = rnorm(n, if_else(state == "idle", 0.8, if_else(state == "stable", 2.2, 3.8)), 0.4),
  sound = rnorm(n, if_else(state == "idle", 45, if_else(state == "stable", 62, 78)), 4),
  torque = rnorm(n, if_else(state == "idle", 10, if_else(state == "stable", 28, 35)), 3)
) |>
  mutate(state = factor(state))

split <- initial_split(dat, prop = 0.8, strata = state)
train <- training(split)
test <- testing(split)
```

## PCA + linear classifier workflow

```{r}
rec <- recipe(state ~ ., data = train) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 2)

model <- logistic_reg(mode = "classification") |>
  set_engine("nnet")

wf <- workflow() |>
  add_recipe(rec) |>
  add_model(model)

fit <- fit(wf, data = train)

pred <- predict(fit, test) |>
  bind_cols(test |> select(state))

metrics <- metric_set(accuracy, kap)(pred, truth = state, estimate = .pred_class)
metrics
