---
title: "04 - HMM Example on Sequential States"
format: pdf
execute:
  echo: true
  warning: false
  message: false
---

## Objective

Estimate hidden operating states from sequential emissions using an HMM.

```{r}
library(tidyverse)
library(HiddenMarkov)
```

## Simulated sequential measurements

```{r}
set.seed(999)

n <- 400
true_states <- numeric(n)
true_states[1] <- 1

for (t in 2:n) {
  if (true_states[t - 1] == 1) {
    true_states[t] <- sample(c(1, 2), 1, prob = c(0.92, 0.08))
  } else {
    true_states[t] <- sample(c(1, 2), 1, prob = c(0.10, 0.90))
  }
}

mu <- c(45, 72)
sd_vec <- c(4, 6)
emission <- rnorm(n, mean = mu[true_states], sd = sd_vec[true_states])
```

## Fit Gaussian HMM and decode states

```{r}
Pi <- matrix(c(0.90, 0.10,
               0.10, 0.90), nrow = 2, byrow = TRUE)

delta <- c(0.5, 0.5)

distn <- "norm"
pm <- list(mean = c(42, 70), sd = c(5, 7))

hmm <- dthmm(x = emission, Pi = Pi, delta = delta, distn = distn, pm = pm)
fit <- BaumWelch(hmm, control = bwcontrol(maxiter = 100))

decoded <- Viterbi(fit)

tibble(
  time = seq_along(emission),
  observed = emission,
  decoded_state = factor(decoded)
)
