---
title: "Multi-agent system for smart machining, part II"
subtitle: "Chapter I-c: exercises"
author: "Paolo Bosetti"
date: "2026-03-20"
date-modified: today
format:
  html:
    code-fold: false
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
engine: knitr
---

# Exercise 1: Data loading and visualization

## Prerequisites

Be sure to havve a look at the preliminary <exercise_01.qmd> to setup your MongoDB instance and load the example data.

## Tasks

### Queries on MongoDB Compass

We want to look at the time series represented by the microphone measurements. Each document in the `acceleration` collection contains 10 samples of the microphone component, each collected every 0.1 ms. We want to expand, or *unwind* these samples to create a single time series of the microphone measurements, with a timestamp for each sample. To do this, we can use the **Aggregation** tab in MongoDB Compass to create a pipeline that performs the following steps:

1. **project**: select only the relevant fields, that is, `message.timestamp` and `message.measurements.microphone`
2. **unwind**: unwind the `message.measurements.microphone` array to create

In practice, you might ask the AI assistant in Compass to write the aggregation pipeline for you, by giving it the following instruction:

> Write an aggregation pipeline that selects the `message.timestamp` and `message.measurements.microphone` fields from the `acceleration` collection, and unwinds the `message.measurements.microphone` array to create a single time series of the microphone measurements, together with the original index of the sample in the 10-sample window.

You should get the following pipeline:

```json
[
  {
    $project: {
      microphone:
        "$message.measurements.microphone",
      timestamp: "$message.timestamp"
    }
  },
  {
    $unwind: {
      path: "$microphone",
      includeArrayIndex: "index"
    }
  }
]
```

You can save this aggregation as a named view `microphone_view`, and then query it as a normal collection to get the microphone time series.

### Queries from R

We start loading some packages:

```{r setup}
#| warning: false
library(tidyverse)
library(mongolite)
library(glue)
options(digits = 15)
options(digits.secs = 3)
```

Now, using the `mongolite` library, write R code to query the `microphone_view`, and load the results in a data frame.

Using the `microphone_view` collection, write a query to select the microphone measurements and their corresponding timestamps, and then create a time series plot of the microphone measurements over time. Using the `timestamp` field (that has a millisecond resolution) and the `index` field (that indicates the position of the sample in the 10-sample window), you can create a new `time` variable that represents the time in seconds from the start of the recording, and use it as the x-axis of the plot.

```{r}
m <- mongo(
  collection = "microphone_view", db = "MASSM_1", 
  url = "mongodb://localhost:27017")
t0 <- "2011-01-01T01:10:34.780+00:00"
query <- glue(
'{{
  "timestamp": {{
    "$gte": {{"$date": "{t0}"}}
  }}
}}'
)

df <- m$find(query=query, limit=300) %>% 
  as_tibble() %>% 
  mutate(time = round(timestamp - min(timestamp) + index * 1e-4, 4), .before = microphone) %>% 
  select(-index)


df %>% 
  ggplot(aes(x = time, y = microphone)) +
  geom_line() +
  labs(x = "Time (s)", y = "Microphone", title = "Microphone time series")
```

::: {.callout-note}
Learn to exploit the aggregation pipeline and the views in MongoDB to simplify your queries and make them more efficient. For example, you can create a view that already computes the `time` variable, so that you can directly query it without having to do any post-processing in R or Python. The AI asistant in MongoDB Compass can help you to write the aggregation pipeline for the view, by giving it the following instruction:

> Write an aggregation pipeline that selects the `message.timestamp` and `message.measurements.microphone` fields from the `acceleration` collection, unwinds the `message.measurements.microphone` array to create a single time series of the microphone measurements, together with the original index of the sample in the 10-sample window, and then creates a new field `time` that represents the time in seconds from the start of the recording, using the `timestamp` field and the `index` field.
:::

### Queries from Python

```{r}
#| echo: false
#| warning: false
library(reticulate)
py_require("pymongo")
py_require("pandas")
py_require("matplotlib")
```

Using the `pymongo` library, write Python code to perform the same queries as above, and load the results in a Pandas data frame. Then, create a plot of the `Fx` component over time, and a histogram of the `microphone` component.

For example, we could run:

```{python}
from pymongo import MongoClient
import pandas as pd
from datetime import datetime
import matplotlib.pyplot as plt
client = MongoClient("mongodb://localhost:27017")
db = client["MASSM_1"]
collection = db["microphone_view"]
t0 = datetime.fromisoformat("2011-01-01T01:08:34.780+00:00")
query = {
    "timestamp": {
        "$gte": t0
    }
}
cursor = collection.find(query).limit(1000)
df = pd.DataFrame(list(cursor))

df["time"] = (df["timestamp"] - df["timestamp"].min()).dt.total_seconds() + df["index"] * 1e-4

plt.plot(df["time"], df["microphone"])
plt.xlabel("Time (s)")
plt.ylabel("Microphone")
plt.show()
```



# Exercise 2: Time series analysis

In R, using the `forecast` package, use `forecasts::ggtsdisplay()` to plot the microphone time series, together with its ACF and PACF. We also split the data into a training set (the first 250 samples) and a test set (the remaining samples), and we plot only the training set to avoid the influence of the test set on the ACF and PACF.

```{r}
#| warning: false
library(forecast)
df <- df %>% mutate(train = row_number() <= 250)
mic_train <- ts(df$microphone[df$train], start = 0, frequency = 10000)
mic <- ts(df$microphone, start = 0, frequency = 10000)
ggtsdisplay(mic_train)
```

Now fit an ARIMA model to the microphone time series, and use it to make predictions for the next 50 samples. Plot the original time series and the predictions together.

```{r}
# Fit ARIMA model
fit <- auto.arima(mic_train, trace=TRUE)

# Forecast next 20 samples
forecasted <- forecast(fit, h=50)

# Plot original time series and predictions
# autoplot(mic, series="Actual") +
ggplot() +
  autolayer(forecasted, series="Forecasted") +
  autolayer(mic, series="Actual") +
  labs(x = "Time (s)", y = "Microphone", title = "Microphone time series with ARIMA predictions")
```

::: {.callout-tip}
Try the same analysis on different time frames. Also try to analyze the ACF and PACF of the time series and fit a custom ARIMA model, instead of using `auto.arima()` (see the `forecast::Arima()` function), to see if you can improve the predictions.
:::


::: {.callout-tip}
Try and find libraries and functions to perform the same analysis in Python. For example, you can use the `statsmodels` library to plot the ACF and PACF, and to fit an ARIMA model. You can also use the `pmdarima` library to perform automatic ARIMA fitting, similar to `auto.arima()` in R.
:::