---
title: "Multi-agent system for smart machining, part II"
subtitle: "Chapter I-a: exercise"
author: "Paolo Bosetti"
date: "2026-03-20"
date-modified: today
format: html
engine: knitr
---

# Exercise 1: Data loading and visualization

## Prerequisites

To run this exercise, you need to have the following tools and libraries installed:

- R and RStudio from [CRAN](https://cran.r-project.org/) and Posit from [Posit](https://posit.co/download/rstudio-desktop/)
- Python 3.x
- [Docker](https://www.docker.com/get-started): install Docker Desktop for your operating system
- [MongoDB Compass](https://www.mongodb.com/products/compass): install the MongoDB GUI client
- a dump of the MongoDB example database (file `mongo_massm_2026-03-20.archive.gz`) from the GDrive shared folder (see Moodle page)

Then you need to start a [MongoDB](https://www.mongodb.com/) instance using Docker:

```bash
mkdir ~/MASSM
docker run -d -p 27017:27017 -v ${PWD}/db:/data/db --name massm-mongo mongo:latest
cat mongo_massm_2026-03-20.archive.gz | docker exec -i massm-mongo mongorestore --gzip --archive --drop
```

or, on Windows PowerShell:

```powershell
mkdir $env:USERPROFILE\MASSM
docker run -d -p 27017:27017 -v "${PWD}\db":/data/db --name massm-mongo mongo:latest
Get-Content mongo_massm_2026-03-20.archive.gz -Raw | docker exec -i massm-mongo mongorestore --gzip --archive --drop
```

Now launch MongoDB Compass and connect to `mongodb://localhost:27017`. You should see a database named `MASSM_1`.

## Database description

The `MASSM_1` database contains two *collections*, `force` and `acceleration`, containing the data acquired during a milling process from a force sensing fixture, a 3-axis accelerometer, and a microphone.

All measurements have been collected at 10kHz rate, originally spanning a total of over 5 million samples, although the force measurements have been downsampled to 200 Hz (providing mean and standard deviation in wondows of 5 ms) to reduce the data size, and the acceleration and sound pressure measurements have been kept at the original 10 kHz rate, but saved in chunks of 10 samples every 1 ms.

Each collection is made by a list of *documents*. Each document is a JSON format object containing relevant information about the acquisition.

For example, the first record in the `force` collection is:

```json
{
  "_id": {
    "$oid": "69bcfdec6d45d8c62c10ef38"
  },
  "timestamp": {
    "$date": "2026-03-20T07:57:32.407Z"
  },
  "message": {
    "agent_id": "",
    "hostname": "Fram-IV-3.local",
    "measurements": {
      "Fx": 0,
      "Fy": 0,
      "Fz": 0,
      "Mz": 0,
      "sd_Fx": 0,
      "sd_Fy": 0,
      "sd_Fz": 0,
      "sd_Mz": 0
    },
    "timestamp": {
      "$date": "2011-01-01T01:06:34.784Z"
    }
  }
}
```

Some notes on the fields:

- `_id`: the unique identifier of the document, automatically generated by MongoDB
- `timestamp`: the time when the document was inserted in the database
- `message.agent_id`: the identifier of the agent that collected the data (empty in this case)
- `message.hostname`: the hostname of the machine where the data was collected
- `message.measurements`: the actual measurements collected, including the mean and standard deviation of the force and torque components
- `message.timestamp`: the actual time when the measurements were collected

Likewise, the first record in the `acceleration` collection is:

```json
{
  "_id": {
    "$oid": "69bd07a76d45d8c62c129cc0"
  },
  "timestamp": {
    "$date": "2026-03-20T08:39:03.970Z"
  },
  "message": {
    "agent_id": "",
    "hostname": "Fram-IV-3.local",
    "measurements": {
      "Ax": [
        -0.87035,
        ... 9 more values ...
      ],
      "Ay": [
        0.35072,
        ... 9 more values ...
      ],
      "Az": [
        -1.24696,
        ... 9 more values ...
      ],
      "microphone": [
        0.20575,
        ... 9 more values ...
      ]
    },
    "timestamp": {
      "$date": "2011-01-01T01:06:34.780Z"
    }
  }
}
```

where this time the `message.measurements` field contains the 10 samples collected in the 1 ms window for each of the 3 acceleration components and for the sound pressure.

# Tasks

## Queries on MongoDB Compass

Using MongoDB Compass, perform the following queries on the `MASSM_1` database:

1. Using the query field and the query option `Project`, select 100 documents from the `force` collection starting 10 seconds after the first timestamp, and only show the `message.timestamp` and `message.measurements.Fx` fields.
2. In the **Aggregation** tab, create a pipeline to compute the mean and standard deviation of the `microphone` component for each document. Save the aggregation as a named View.

::: {.callout-tip}
Liberally use the AI assistant: learn how to give it precise and clear instructions, and ask it to write the query for you.
:::

::: {.callout-important}
Aggregations are powerful tools that can reduce computational load on database clients that consume data for analytics.

Note that with the **green save button** you can save the aggregation in the collection, so that you can run it again later, or you can **create a view**. A view is a virtual collection that is not materialized in the database, but it can be queried as a normal **read-only collection**, and it will execute the aggregation pipeline on the fly. This is useful to create custom data transformations that can be reused in different contexts.
:::

Once you have customized your data query or transformation, you can export the results as a CSV file, and then load it in R or Python for further analysis and visualization. This approach, though, **is not ideal for large datasets** (you are duplicating the data!), and it is better to use a database client library to query the database directly from your code, as we will see in the next exercises.

## Queries from R

We start loading some packages:

```{r}
#| warning: false
library(tidyverse)
library(mongolite)
library(glue)
options(digits = 15)
options(digits.secs = 3)
```

::: {.callout-warning}
If your R installation does not have the above packages, you can install them using the command `install.packages()` function.
:::

Now, using the `mongolite` library, write R code to perform the same queries as above, and load the results in a data frame. Then, create a plot of the `Fx` component over time, and a histogram of the `microphone` component.

For example, we could run:

```{r}
m <- mongo(collection = "acceleration", db = "MASSM_1", url = "mongodb://localhost:27017")
t0 <- "2011-01-01T01:07:34.780+00:00"
query <- glue(
'{{
  "message.timestamp": {{
    "$gte": {{"$date": "{t0}"}}
  }}
}}'
)
fields <- '{
  "_id": 0, 
  "message.timestamp": 1,
  "message.measurements.microphone": 1
}'
df <- m$find(query=query, fields = fields, limit=200) %>% as_tibble()
tibble(
  timestamp = df$message$timestamp,
  microphone = map_dbl(df$message$measurements$microphone, mean)
) %>% 
  ggplot(aes(x=timestamp, y=microphone)) +
  geom_line() +
  scale_x_datetime(date_labels = "%M.%OS", date_breaks = "0.05 sec")
```

::: {.callout-tip}
The query result is a rather complicate data frame, for its columns are nested lists. Try to use an aggregation of a saved View to make it simpler and more efficient to load in R.
:::

## Queries from Python

```{r}
#| echo: false
#| warning: false
library(reticulate)
py_require("pymongo")
py_require("pandas")
py_require("matplotlib")
```

Using the `pymongo` library, write Python code to perform the same queries as above, and load the results in a Pandas data frame. Then, create a plot of the `Fx` component over time, and a histogram of the `microphone` component.

For example, we could run:

```{python}
from pymongo import MongoClient
import pandas as pd
from datetime import datetime
import matplotlib.pyplot as plt
client = MongoClient("mongodb://localhost:27017")
db = client["MASSM_1"]
collection = db["acceleration"]
t0 = datetime.fromisoformat("2011-01-01T01:07:34.780+00:00")
query = {
    "message.timestamp": {
        "$gte": t0
    }
}
fields = {
    "_id": 0,
    "message.timestamp": 1,
    "message.measurements.microphone": 1
}
cursor = collection.find(query, fields).limit(100)
df = pd.DataFrame(list(cursor))
df["microphone"] = df["message"].apply(lambda x: sum(x["measurements"]["microphone"]) / len(x["measurements"]["microphone"]))
df["timestamp"] = df["message"].apply(lambda x: x["timestamp"])
plt.plot(df["timestamp"], df["microphone"])
plt.xlabel("Time")
plt.ylabel("Microphone")
plt.show()
```

::: {.callout-tip}
As said above for R, the query result is a rather complicate data frame, for its columns are nested lists. Try to use an aggregation of a saved View to make it simpler and more efficient to load in Python.
:::